Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-11-23 edubots_surveys.Rnw
Corrected a typo and grammar

Given the potential benefits that chatbots could bring to the classroom, the
focus of the
% EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
{\sf anonymized European Project}, the sponsor of this study, is to
\textcolor{blue}{explore best practices and innovative use of chatbots, and to create 
a learning community of educators in higher education institutions.
The majority of studies conducted so far have studied how these 
messaging applications and/or chatbots are used to deliver personalised
learning in classrooms that occurs anytime anywhere, 
promote collaborative learning experiences, group discussions \cite{panah2020study}
and for boosting students' sense of belonging to their 
institutions \cite{abbas2021onlinechat}. Hence, due to the paucity of research
that explores the use of these technologies from the teachers' perspectives,
this paper aims to investigate their opinions/needs and the
challenges/opportunities of adopting these technologies in classrooms
to paint a better picture of how they can positively contribute to enhancing
the learning process in higher education institutions.}

@@

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed\textcolor{blue}{; this will be explained more fully in the corresponding section}. The responses' analysis
entailed the creation of a second survey to which 200 teachers' responses were
collected\textcolor{blue}{, with the methodology and collective characteristics that will be explained below. The questions of the second survey focussed on discussing teachers'
needs, opinions and preferences of the development of future technology-enhanced
tools and its potential impact on educational institutions’ policies. In order to assess these, this paper investigated the following research questions}:

@@

\textcolor{blue}{This study aims to collect rich data about the use of messaging applications and chatbots
by educators in both universities and colleges\footnote{In Spain, they are called simply high schools, and deliver
tertiary or associated degrees that do not have the consideration of university degrees}.
To achieve this goal, an online survey was designed and developed using Google Forms
in Spanish}. Initially, the intention was to mirror the questions used in
\cite{MoraChatbots2021:anon}; which explored students' views about the use of messaging applications in classrooms.
However, the initial exploration of results led us
to create a second survey in order to collect more data about educators' preferences in particular. The \textcolor{blue}{new survey} items 
focused on the use of messaging apps in the teaching practice, the type of chatbot use cases that teachers consider
useful to be used in their teaching, and the type of interactions that teachers
prefer to have with their students while using messaging apps in their teaching.

The {\bf initial survey} was done in two separate Google forms, one for
university and other for tertiary non-university teachers. Questions and
responses were the same, except for the type of tertiary degrees that
\textcolor{blue}{were considered. The survey was piloted by the authors of the paper and their colleagues before using it in the 
study.  All the comments collected from this trial were incorporated in the survey.
We used mainly mailing lists, as well as Telegram groups,
to post the survey link to reach out to more educators}. The university form was sent to Spanish
(mainly Andaluc\'{\i}a and Galicia) universities, and also  universities in Costa Rica and
Mexico. The tertiary (non-university/college) teachers were based mainly in
Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of
2021, post-pandemic and while, at least in Spain, many universities had
mandatory virtual teaching.

@@

% Fran: I think figures with data about the use of messaging apps should come here. Then, we say we also asked about changes after covid and include data about that
Results in Figs. \ref{figure:provided*chatbots_use},
\ref{figure:whatsapp*chatbots_use}, \ref{figure:telegram*chatbots_use} do not
show significant differences in the use of instant messaging apps between
teachers from University and vocational education. In general, most teachers
prefer messaging apps provided by their own institutions (we also show the
results for the two most popular messaging app platforms: WhatsApp and
Telegram). With respect to specific disciplines, Engineering and Technology
teachers are more active in their use, \textcolor{blue}{but the amount of teachers from Humanities 
who answered they used these apps in their classes is also remarkable}
(around 60\% use the apps provided by their institutions).

@@

A total of 205 teachers responded to the second survey, 124 males (60.5\%), 65 females (31.7\%) and 16 individuals who chose not to identify their gender (7.8\%). The majority of the teachers were aged 45-55 years old (70; 34.1\%), followed by individuals aged 35-45 (67, 32.7\%), and 25-35 (42; 20.5\%), while the minority was aged $>$~55 years old (26; 12.7\%). In terms of teaching experience, 59 teachers (28.8\%) had an experience of 16-25 years, followed by teachers who had a teaching experience of 06-15 years (51; 24.9\%), 0-5 years (51; 24.9\%) and more than 25 years of experience (44; 21.5\%). The teachers were asked about what kind of interaction media features they value the most; the frequencies and percentages of their responses are given in Table \ref{tab:interaction}.

@@

Analytics & 108 & 52.7 & 97 & 47.3 \\
Connectivity & 119 & 58.0 & 86 & 42.0 \\
Familiarity & 121 & 59.0 & 84 & 41.0 \\
Hidden Phone Number & 113 & 55.1 & 92 & 44.9 \\
Horizontally & 134 & 65.4 & 71 & 34.6 \\
Official formation & 42 & 20.5 & 163 & 79.5 \\
Pluggability & 65 & 31.7 & 140 & 68.3 \\
Sustainability & 157 & 76.6 & 48 & 23.4\\
Unidirectionality & 27 & 13.2 & 178 & 86.8 \\
Officiality & 150 & 73.2 & 55 & 26.8 \\
Synchrony & 45 & 22.0 & 160 & 78.0\\
Other & 5 & 2.4 & 200 & 97.6 \\
\hline
\end{tabular}
\vspace{0.8em}
\caption{Interaction media features valued by teachers. Details about each feature can be read in Section \ref{sec:meth}.}
\label{tab:interaction}
\end{table}

@@

The key findings of this study shed the light on educators’ preference to use
messaging applications that their institutions support.  Technology adoption
literature often \textcolor{blue}{focuses} on users’ perceptions of the technology’s usefulness and
ease of use as important prerequisites for successful adoption and
utilization. Nevertheless, in higher education, institutions’ role in
integrating these tools to their educational systems can improve the uptake of
these applications and shape the social and educational experiences of their
students. \textcolor{blue}{To accomplish this fruitful integration}, institutions have to ensure that
these messaging applications not only are GDPR compliant to keep students’ data
secure but also should provide IT support to all stakeholders who use the
applications.

@@

% Including questions in surveys
The {\bf first survey}, which \textcolor{blue}{focused on} the use of messaging apps in class, and which tried to address mainly RQ2 and RQ3, 
% Antonio - I guess it refers to RQ3. ;)
includes the following questions:

@@

\begin{figure}[h!tbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/survey-2.png}
\end{center}
    \caption{\textcolor{blue}{Screen capture of the Spanish-language form used for the survey}}
\label{figure:survey2}
\end{figure}
The {\bf second survey} followed the same paths, only in this case there was a single questionnaire, and with a different scope, focusing on the kind of interaction media features that teachers value the most. \textcolor{blue}{A capture of these questions is included in Figure \ref{fig:survey2}}
% Olia - see my comment above.
It was filled by teachers that were also students in a Professional Development course (on the use of new technologies in higher education, around 1/4 of them) as well as University of Granada teachers who knew about it by emails from the authors, or from the vicedeanship for International Relations that included it in its newsletter. This means that there might be a higher proportion of 1) teachers with few years of experience 2) teachers from the computer science faculty and 3) teachers from Granada. We think, however that there's no explicit bias in this selection, although of course specific percentages will vary. These questions tried to address mainly RQ1 and RQ4.

>>>>

Given the potential benefits that chatbots could bring to the classroom, the
focus of the
% EDUBOTS KA2 European project \footnote{https://www.edubots.eu},
{\sf anonymized European Project}, the sponsor of this study, is to
\textcolor{blue}{explore best practices and innovative use of chatbots, and to create 
a learning community of educators in higher education institutions.
The majority of studies conducted so far have studied how these 
messaging applications and/or chatbots are used to deliver personalised
learning in classrooms that occurs anytime anywhere, to
promote collaborative learning experiences, group discussions \cite{panah2020study}
and to boost students' sense of belonging to their institutions \cite{abbas2021onlinechat}. Hence, due to the paucity of research
that explores the use of these technologies from the teachers' perspectives,
this paper aims to investigate their opinions/needs and the
challenges/opportunities of adopting these technologies in classrooms
to paint a better picture of how they can positively contribute to enhancing
the learning process in higher education institutions.}

@@

Initially, a single survey was developed for this purpose and responses from 300
higher education teachers were collected and analysed\textcolor{blue}{; this will be explained more fully in the methodology section}. The responses' analysis
entailed the creation of a second survey to which 200 teachers' responses were
collected\textcolor{blue}{, with the methodology and collective characteristics that will be explained below. The questions of the second survey focussed on discussing teachers'
needs, opinions and preferences of the development of future technology-enhanced
tools and its potential impact on educational institutions’ policies. In order to assess these, this paper investigated the following research questions}:

@@

\textcolor{blue}{This study followed a quantitative approach towards addressing its research objective, by collecting rich data about the use of messaging applications and chatbots
by educators in both universities and colleges\footnote{In Spain, they are called simply high schools, and deliver
tertiary or associated degrees that do not have the consideration of university degrees}.
To achieve this goal two online surveys were designed and developed using Google Forms
in Spanish. They are comprised of six questions focusing on demographic data
(i.e., sector, gender, degrees, discipline, age, teaching experience)
followed by several multiple choice questions allowing the participants to choose multiple answers.}

\textcolor{blue}{In the first survey, the multiple choice questions are
focused on the use of messaging apps in the teaching practice, the type of chatbot use cases that teachers consider
useful to be used in their teaching and the impact of COVID on the teaching practice; please check them out to the full extent in the Appendix.\footnote{This {\bf first survey} was made in two separate Google forms, one for
university and other for tertiary non-university teachers. Questions and
responses were the same, except for the type of tertiary degrees that
were considered.}}

\textcolor{blue}{This survey was piloted by the authors of the paper and their colleagues before using it in the 
study.  All the comments collected from this trial were incorporated in the survey.
We used mainly mailing lists, as well as Telegram groups,
to post the survey link to reach out to more educators. The university form was sent to Spanish
(mainly Andaluc\'{\i}a and Galicia) universities, and also  universities in Costa Rica and
Mexico. The tertiary (non-university/college) teachers were based mainly in
Andaluc\'{\i}a. Dissemination and answers took place in the first quarter of
2021, post-pandemic and while, at least in Spain, many universities had
mandatory virtual teaching.}

@@

% Fran: I think figures with data about the use of messaging apps should come here. Then, we say we also asked about changes after covid and include data about that
Results in Figs. \ref{figure:provided*chatbots_use},
\ref{figure:whatsapp*chatbots_use}, \ref{figure:telegram*chatbots_use} do not
show significant differences in the use of instant messaging apps between
teachers from University and vocational education. In general, most teachers
prefer messaging apps provided by their own institutions (we also show the
results for the two most popular messaging app platforms: WhatsApp and
Telegram). With respect to specific disciplines, Engineering and Technology
teachers are more active in their use, \textcolor{blue}{but the number of teachers from Humanities 
who answered they used these apps in their classes is also remarkable}
(around 60\% use the apps provided by their institutions).

@@

A total of 205 teachers responded to the second survey, 124 males (60.5\%), 65 females (31.7\%) and 16 individuals who chose not to identify their gender (7.8\%). The majority of the teachers were aged 45-55 years old (70; 34.1\%), followed by individuals aged 35-45 (67, 32.7\%), and 25-35 (42; 20.5\%), while the minority was aged $>$~55 years old (26; 12.7\%). In terms of teaching experience, 59 teachers (28.8\%) had an experience of 16-25 years, followed by teachers who had a teaching experience of 06-15 years (51; 24.9\%), 0-5 years (51; 24.9\%) and more than 25 years of experience (44; 21.5\%). The teachers were asked about what kind of interaction media features they value the most \textcolor{blue}{; i.e. what are their preferred utilities in the applications/tools they use to interact with their students.} The frequencies and percentages of their responses are given in Table \ref{tab:interaction}.

@@

Analytics & 108 & 52.7 & 97 & 47.3 \\
Connectivity & 119 & 58.0 & 86 & 42.0 \\
Familiarity & 121 & 59.0 & 84 & 41.0 \\
Hidden Phone Number & 113 & 55.1 & 92 & 44.9 \\
Horizontally & 134 & 65.4 & 71 & 34.6 \\
Official formation & 42 & 20.5 & 163 & 79.5 \\
Pluggability & 65 & 31.7 & 140 & 68.3 \\
Sustainability & 157 & 76.6 & 48 & 23.4\\
Unidirectionality & 27 & 13.2 & 178 & 86.8 \\
Officiality & 150 & 73.2 & 55 & 26.8 \\
Synchrony & 45 & 22.0 & 160 & 78.0\\
Other & 5 & 2.4 & 200 & 97.6 \\
\hline
\end{tabular}
\vspace{0.8em}
\caption{Interaction media features valued by teachers. \textcolor{blue}{Details about each of these features can be read in Appendix I (Section \ref{app:questions}, second survey)}.}
\label{tab:interaction}
\end{table}

\textcolor{blue}{As it can be seen in the table, the main responses aim to use a sustainable and official application, i.e. both being approved or provided by the University (or the educational centre), and also maintained by the technical staff of the centre instead of giving this responsability/task to the teacher.
In addition, it is very important for the teachers that the used tools allow to communicate all the members of the classroom, including themselves; as well as being already known or used in everyday tasks, e.g. Telegram or WhatsApp.}

@@

The key findings of this study shed the light on educators’ preference to use
messaging applications that their institutions support.  Technology adoption
literature often \textcolor{blue}{focuses} on users’ perceptions of the technology’s usefulness and
ease of use as important prerequisites for successful adoption and
utilization. Nevertheless, in higher education, institutions’ role in
integrating these tools to their educational systems can improve the uptake of
these applications and shape the social and educational experiences of their
students. \textcolor{blue}{To accomplish this fruitful integration}, institutions have to ensure that
these messaging applications not only are GDPR compliant to keep students’ data
secure, but also should provide IT support to all stakeholders who use the
applications.

@@

\begin{figure}[h!tbp]
\begin{center}
\includegraphics[width=0.85\textwidth]{figures/survey-1.png}
\end{center}
    \caption{\textcolor{blue}{Screen capture of the Spanish-language form used for the first survey}}
\label{figure:survey1}
\end{figure}
% Including questions in surveys
The {\bf first survey}, shown in its original form in Figure \ref{figure:survey1}, which \textcolor{blue}{focused on} the use of messaging apps in class, and which tried to address mainly RQ2 and RQ3, 
% Antonio - I guess it refers to RQ3. ;)
includes the following questions:

@@

\begin{figure}[h!tbp]
\begin{center}
\includegraphics[width=\textwidth]{figures/survey-2.png}
\end{center}
    \caption{\textcolor{blue}{Screen capture of the Spanish-language form used for the survey}}
\label{figure:survey2}
\end{figure}
The {\bf second survey}, shown in its original form in Figure \ref{figure:survey2}, followed the same paths, only in this case there was a single questionnaire, and with a different scope, focusing on the kind of interaction media features that teachers value the most. \textcolor{blue}{A capture of these questions is included in Figure \ref{fig:survey2}}
% Olia - see my comment above.
It was filled by teachers that were also students in a Professional Development course (on the use of new technologies in higher education, around 1/4 of them) as well as University of Granada teachers who knew about it by emails from the authors, or from the vicedeanship for International Relations that included it in its newsletter. This means that there might be a higher proportion of 1) teachers with few years of experience 2) teachers from the computer science faculty and 3) teachers from Granada. We think, however that there's no explicit bias in this selection, although of course specific percentages will vary. These questions tried to address mainly RQ1 and RQ4.
